---
title: "Analyzing Determinants of Crime"
author: "Rowan Cassius, Michael Steckler, Julian Pelzner"
date: "12/3/2019"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This study was conducted for the reelection campaign of the Governor of North Carolina. Political consultants have identified crime as a primary voter initiative. In this report, we provide an analysis that illuminates the determinants of crime, and we suggest policy actions that can be implemented at the local government level. Our key research question is: How does a county’s living conditions and likelihood of penalization affect its crime rate? To that end, our outcome variable of interest is crime rate per county. Our key explanatory variables of interest are density and the probability of arrest. 

This report aims to contriute to the discussion through an analysis of crime rates of counties in North Carolina in 1987. The data we have analyzed contains information about crime rates and various socioeconomic, geographic and demographic factors at the county level for 91 of North Carolina's 100 counties.

Through our analysis we seek to understand how effective increased penalization is at combating crime and how large an effect a county's physical environment has on the amount of crime it experiences.

## Data Cleaning

```{r, echo = FALSE}
# Dependencies
library("stargazer")
library("car")
library("ggplot2")
library("gridExtra")

# Loading Data
crime <- read.csv('crime_v2.csv')
```

```{r, echo=FALSE, results='hide'}
# Exploration:
# Missing value summary
summary(is.na(crime))
# Examining missing Values
crime[91:nrow(crime), ]
paste0("Number of missing values in first 92 rows: ", sum(is.na(crime[1:91,])))
# Variales Types Summary
str(crime)
```


```{r, echo=FALSE, results='hide'}
# Cleaning changes:

# Omitting NAs
crime <- na.omit(crime)

# Casting factor as numeric
crime$prbconv <- as.numeric(crime$prbconv)

# Removing year and county identifiers
crime <- crime[!names(crime) %in% c('year', 'county')]

# Summary
summary(crime)

# Examining two of the probaility variables
print('Probaility of arrest summary:')
summary(crime$prbarr)
print('Probability of conviction summary:')
summary(crime$prbconv)

# Casting proability of conviction as probability
crime$prbconv <- 1.0/100 * crime$prbconv
```

```{r, fig.align='center', fig.height=2.75, fig.width=4, fig.cap="Distribution of Probility of Arrest", echo=TRUE}
# Examining outlier in probability of arrest.
#hist(crime$prbarr)
ggplot(crime, aes(x=prbarr)) + 
  geom_histogram(bins = 20, fill = "dodgerblue4") +
  xlab("Probability of Arrest") +
  ylab("Count")
```
By looking at the histogram of `prbarr`, the variables has one outlier at `r round(max(crime$prbarr), 2)`. This is puzzling as it suggest that the number of convictions in the associated county was greater than the number of arrests in 1987. Some plausible explantions for this apparent outlier include the following:

* The number of convictions includes spillover from pervious years. That is, the number of convictions includes convictions from 1986 and before while the number of arrests only includes arrests from 1987.

* The number of convictions includes spillover from neighoring counties. That is, some criminals were arrested in neighboring counties and tried in this county of interest, thus letting the apparent number of convictions exceed the apparent number of arrests.

* The outlier is a mistranscription. The official who reported the figure meant to report `r 1 - round(max(crime$prbarr), 2)`.

While plausile, it is not safe to assume the outlier is a mistranscription, so we most stongly speculate that it is a result of spill over either from a different year or a different county, or both. All of the cases of spillover suggest that the observatio containing this outlier did not come from the underlying population we intend to study: counties which each report their own crime and crime-related information exclusively in the year of 1987. For this reason, we removed the observation containing this ourlier from the analysis.

```{r, echo=FALSE}
crime <- crime[crime$prbarr <= 1.0, ]
```

After these changes, the data are vetted and ready for analysis.

## Exploratory Analysis

In this analysis, we seek to understand the extent to which living conditions and the penalization of criminality are related to crime. It is necessary to choose variables representing all of the concepts we wish to study. Among all the variables in the dataset, we select `crmrte`, which represents the number of crimes committed per person in a given county, as the dependent variable we use to measure crime. 

We also select `prbarr`, which represents the ratio of offenses to arrests, also known as the probability of arrest, as a proxy for measuring the county's efforts to penalize crime on the street level. We hypothesize that an increased likelihood of arrest stifles crime rates. If criminals are less likely to be arrested after are commiting offenses, we argue that this will make them less inclined to commit the offense in the very first place. In addition,  a higher probability of arrest could indicate that the police are more effective at catching people who commit crimes in the county. While it is not a perfect proxy for penalization, probability of arrest provides a basic metric to evaluate how likely arrests are to be made for crimes committed in a county.

Our primary explanatory variable to proxy living coniditions is the `density` variable, which represents the the number of people per square mile in a county. We hypothesize that denser areas tend to make people more irritable due to increased economic competition and lack of comfortability. Moreover, it can be argued that there are more opportunities to commit crimes in denser areas. In summary, the key explanatory variables in our baseline model will be probability of arrest and density. The next figure shows the marginal distributions of both key explanatory variables and of crime rate.


```{r, fig.align='center', fig.height=3, fig.width=9, fig.cap="Marginal Distributions of Crime Rate, Probability of Arrest and Density", echo=TRUE}
# Crime histogram
plot.crime = ggplot(crime, aes(x=crmrte)) + 
  geom_histogram(bins = 20, fill = "red3") +
  xlab("Crime Rate") +
  ylab("Count") +
  ggtitle("Crime Rate")

# Arrest Probability histogram
plot.arrest = ggplot(crime, aes(x=prbarr)) + 
  geom_histogram(bins = 20, fill = "dodgerblue4") +
  xlab("Probability of Arrest") +
  ylab("") +
  ggtitle("Probability of Arrest")

# Density histogram
plot.density = ggplot(crime, aes(x=density)) + 
  geom_histogram(bins = 20, fill = "seagreen4") +
  xlab("Density") +
  ylab("") +
  ggtitle("Density")

grid.arrange(plot.crime, plot.arrest, plot.density, nrow = 1, ncol = 3)
```

Each distribution has some right skew and a couple outliers, but this is to be expected of complex data.


```{r, fig.height=7, fig.width=7, fig.cap="Scatter Plots of Crime against Probability of Arrest and Density", echo=FALSE}

p1 <- ggplot(crime, aes(x=prbarr, y=crmrte)) +
  geom_point() +
  geom_smooth(method="gam", color = "dodgerblue4", fill = "dodgerblue4") +
  xlab("Probability of Arrest") +
  ylab("Crime Rate") +
  ggtitle("Crime Rate and Probability of Arrest")

p2 <- ggplot(crime, aes(x=density, y=crmrte)) + 
  geom_point() +
  geom_smooth(method="gam", color = "seagreen4", fill = "seagreen4") +
  xlab("Density") +
  ylab("Crime Rate") +
  ggtitle("Crime Rate and Density")

p3 <- ggplot(crime, aes(x=prbarr, y=log(crmrte))) +
  geom_point() +
  geom_smooth(method="gam", color = "dodgerblue4", fill = "dodgerblue4") +
  xlab("Probability of Arrest") +
  ylab("Log-Crime Rate") +
  ggtitle("Log-Crime Rate and Probability of Arrest")

p4 <- ggplot(crime, aes(x=density, y=log(crmrte))) + 
  geom_point() +
  geom_smooth(method="gam", color = "seagreen4", fill = "seagreen4") +
  xlab("Density") +
  ylab("Log-Crime Rate") +
  ggtitle("Log-Crime Rate and Density")

grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

The scatter plots above compare the scatter plots of crime rate with each primary explanatory variable and the scatter plots of log-crime rate with the same variables. It appears that the untransformed specifications have the strongest linear relationship with crime rate, so we will specifiy our baseline model without transforming crime rate. Refraining from transforming the response variable will enable us to directly interpret the model coefficients as the associated changes in crime rate for a single unit increase in each explanatory variable.

The first scatter plot illustrates that likelihood of arrest has a negative correlation with crime rate, supporting our hypothesis that increasing crackdown on crime is associated with a lower crime rate. The second plot supports our second hypothesis that increases in population density are associated with increased crime rates. 

## Baseline Model Specification. 

$$
crimrate = \beta_0 + \beta_1prbarr + \beta_2density + \epsilon
$$
```{r}
model.1 <- lm(crmrte ~ prbarr + density, data=crime)
```

### Assessing CLM Assumptions

1. **Linearity**

* After our bivariate exploratory analyses of crime rate with probability of arrest and crime rate with density, we chose the specification offering relationships that appear linear. However, we can argue that the model is linear no matter what because we have not yet contrained $\epsilon$.

2. **Random Sampling**

* We can infer that because there are 97 data points, out of 100 possible counties in the state, that the researchers intended to complete a census of the population of counties in the state. However, because data from a few counties was inaccessible, they limited the study to a convenience sample of the remaining counties. Since over 5% of the population was sampled, we can not reasonably argue that the data was collected from a random sample. Nonetheless, with over 90% of counties in the state represented, we can conclude that the results of our models will offer useful insights into the causal factors behind crime rate on a statewide level. We assume that the inavailability of data in some counties is not connected to the crime rates therein in any significant way.

3. **Multicollinearity** 

* To test for multicollinearity, we have calculated the inflated variance factors for all of the regressors in the first model. The variance inflation factors for both probability of arrest and density are close to 1, so there is no problematic multicollinearity and, therefore, no perfect multicollinearity.

```{r}
round(car::vif(model.1), 2)
```

4. **Zero Conditional Mean**

* To examine the zero conditional mean assumption, we have plotted the residuals as a function of the fitted values. By inspection of the residuals vs fitted values plot, there is not substantial evidence of violation of the zero conditional mean assumption. The conditional mean remains close to zero but tends downward slightly as the fitted values increase.

```{r, fig.height=2.5, fig.width=3, fig.cap="Baseline Models Residuals vs Fitted Values Plot", echo=FALSE}
ggplot(crime, aes(x=model.1$fitted.values, y=model.1$residuals)) +
  geom_point() +
  geom_smooth() +
  xlab("Fitted Values") +
  ylab("Residuals") +
  ggtitle("Residuals vs Fitted Values")
```

5. **Heteroscadascity**

* By examination of the residuals vs fitted values plot, there appears to some degree of heteroscedasticity because the magnitude of the residuals taper off at both the highest and lowest fitted values. To address this, we will use heteroscedasticty robust standard errors when making inferences.

6. **Normality of Errors**

* By inspection of the residuals' normal Q-Q plot and their distribution, the error distribution suffers from a fat right tail and a short left tail. While there this means the error distribution deviates from normality somewhat, the sample size of `r nrow(crime)`, which is far greater than 30, enables us to confidently invoke the central limit theorem and conclude that in spite of the errors' non-normality, we still approximate the sampling distributions of the model's coefficients.

```{r, fig.height=3.5, fig.width=7, fig.cap="Residuals' distribution and Normal Q-Q Plot", echo=TRUE}
residual <- data.frame(model.1$residuals)
p.resid <- ggplot(residual, 
                  aes(x=model.1.residuals)) +
  geom_histogram(bins=20) +
  xlab("Residual") +
  ggtitle("Residual Histogram")
q.resid <- ggplot(residual, aes(sample = model.1.residuals)) +
  stat_qq() + 
  stat_qq_line() +
  ggtitle("Residual Q-Q Plot")

grid.arrange(p.resid, q.resid, nrow = 1, ncol = 2)

```
### Baseline Model Results






